Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-09-08 TACO-paper-v2/TACO-Rebuttal.txt

We thank our reviewers (R1-R4). We appreciate the positive feedback and the
critical comments on our paper. Below we answer the various points arose by the
reviewers.

R1:
We agree the big missing part was a discussion on the generality of our
approach and the applicability of the code transformations to other
computational domains (or, perhaps, their integration within a general-purpose
compilers). We added therefore a complete new section (Section 6) that explains
these aspects in details. We believe this addition was necessary to let readers
understand how to transfer our work to a different domain or compiler (or, at
least, to let them gain something from our experience of designing/developing 
the code transformations and COFFEE).

We also addressed all of the other useful comments (1-8) provided:
1) The goal is to develop a class of optimizations for an infinite variety of 
local assembly codes. In that paragraph we say "(...) makes it hard to 
determine a single or specific sequence of transformations that is successfully 
applicable to all problems (...)", which implies that different finite element 
problems may require different code optimizations due to the inherent diversity 
in mathematical expressions and loop trip counts (this aspect is reiterated 
throughout the paper). Therefore, a compiler-based approach, in which the input 
and some parameters are analyzed to decide how to optimally transform the code 
is needed to maximize the performance. Another possibility would be a 
library-based approach, although it would be quite pointless since we can 
generate the most suitable code for each problem.
2) The variation in loop trip counts impacts loops transformation. If the loop 
is relatively large, for example, vector-register tiling is likely to be useful 
to minimize register spilling; if it is small, then full or partial unrolling 
is an option. Understanding the border between "small" and "large" is 
challenging.  The cost model helps in this direction. We tried to clarify the 
point you make in various sections in the paper: the sentence at page 6 has 
been re-written to anticipate the fact that "different loop sizes require 
different transformations". Then we elaborate on the importance and the impact 
of loop unrolling in Section 4 (there is a complete new paragraph about that).  
Finally, in Section 5.2.5, we demonstrate our claim when discussing the impact 
of vector-register tiling (paragraphs 4, 5 and 6 of that section) in our 
benchmarks.
3) Clarified in the paper by adding/re-writing a few sentences. The short 
answer is: yes.
4) Explained in the paper.
5) We agree. We removed the code listing and expanded Figure 4.
6) Explained in the paper.
7) This was already explained by the sentence "Other optimization levels 
performed, in general, slightly worse."
8) We believe each of those tables can be represented by 6 different charts (1 
for each <Architecture, Equation> examined>). We have not done that for lack of 
space (the submission for acceptance is limited to 25 pages) but we can easily 
do that, if really needed, in the final version of the paper (which can be 
30-page long)

@@

******************************************************************************

R2:
Two main critiques are moved to our paper.

a) We do not think that "the techniques of the paper are mostly interesting to 
a subset of the people in the intersection of code optimization and finite 
element methods". However, we do agree that the paper was missing an in-depth 
explanation about how to re-adapt/re-use our transformations in other contexts, 
perhaps other DSLs or even general-purpose compilers. Also, we believe that our 
experience of developing COFFEE and integrating it in a real-world framework 
based on a domain-specific language can be of inspiration for other research 
and/or engineering work. We therefore introduced a complete new section, 
Section 6, about the generality of our approach and the applicability of the 
code transformations to other computational domains. We hope you can read this 
section carefully, because it represents novel contribution of the paper and 
aims to answer exactly the concern you pointed out.

b) The results of the paper are *really* applicable to other kernels beyond 
those analyzed and presented. First of all, we want to emphasize a number of 
aspects. What is important to clarify is that kernels are not "mere small 
pieces of code", but rather, in general, quite complex numerical routines 
evaluating integrals in an
element of the discretized equation domain, produced by means of real (used by 
many scientists around the world) domain-specific language and compiler.  There 
is an infinite variety of kernels that can be generated in Firedrake; the 
generated code (e.g. loop sizes and mathematical expression) depends on the 
partial differential equation being solved. This is explained and re-iterated 
in the paper. Also, varying the polynomial order of the method or the element 
shape (e.g. triangles, tetrahedrons, prisms), as we do and explain in the 
performance evaluation section, changes the generated code by increasing loops 
and arrays sizes and by making mathematical expressions more articulated (this 
happens, for example, when moving from 2D to 3D domains). So we are already 
testing many different code variants (three different kernels, and then for 
each kernel we try multiple polynomial orders and element shapes). This is 
carefully explained in Section 5.2.1. Now, premised this, back to your point, 
which is "applicability and effectiveness of our transformations in other local 
assembly kernels". Our answer is twofold:

- Section 5.2.1 explains in-depth the motivations behind the choice of the 
  three problems (Helmholtz, Diffusion, Burgers), explaining that "(...) 
distinct problems, possibly arising in completely different fields, may employ 
(subsets of) the same differential operators of our benchmarks, which implies 
similarities and redundant patterns in the generated code.  Consequently, the 
proposed code transformations have a domain of applicability that goes far 
beyond that of the three analyzed equations. (...)". Then we re-inforce this 
later on in the section and in the paper. Other kernels arising from completely 
different equations will still have the same structure of the kernels we 
examined (e.g. same memory access pattern, same ranges of loop sizes, same 
problems concerning the presence of loop-invariant code and register pressure) 
as well as similar patterns in the mathematical expressions, so there is no 
evident reason for which our code optimizations would not be effective. The 
critical issue in our research was to understand what transformations our local 
assembly kernels required, and how to instantiate parametric transformations 
(vector-register tiling, loop interchange, expression splitting) optimally in 
each possible problem (this led to the development of the cost model).

- In Section 5.3 we provide a full-application study in which we analyze a 
  different equation (Linear Elasticity), which brings the total number of 
studied equations to 4 (Helmholtz, Diffusion, Burgers, Elasticity)
Adding more kernels to the performance evaluation section would make the paper 
only more chaotic: we selected these three problems to highlight all pros and 
cons of our code transformations. This is the reason for which we comment on 
the performance of the code transformations (and their interplay) in specific 
sub-sections, referring to, describing, and carefully analyzing the provided 
plots.

Other concerns:
- It is actually not true that we do not consider other optimization 
  strategies: 1) a comparison with using linear algebra libriries (the 
comparison uses hand-crafted kernels) and 2) a comparison with the FEniCS Form 
Compiler's built-in optimization system (which is the result of a cited paper) 
are provided.
- We agree it is useful to report speed ups and execution times for the whole 
  finite element calculation. We have added a new section (Section 5.3) about a 
full-application study, showing exactly these values.
- In some points in the paper (introduction and related work) we cite studies 
  concerning local assembly and GPUs. These works however deeply differ from 
ours for a number of reasons, including the facts that they do not rely on 
automated code generation, they are limited to only very low-order methods 
(typically only polynomial order 1), and they do not try to optimize for 
cross-loop arithmetic intensity as we do. The goal in these works is rather to 
understand how to effectively parallelize the local assembly code (so as to 
have multiple threads executing a single kernel, as opposed to our execution 
model in which we assume a single thread per kernel) in specific equations (not 
in a possibly infinite range as we do). A critical issue when executing local 
assembly kernels on GPUs is represented by memory requirements. We give here 
two examples of this problem, although more issues could be found. First, the 
transfer of the mesh from host to GPU may be a notable obstacle: if too big, 
the mesh
will not fit the available GPU memory, and distributed execution, which is 
marginally supported in available finite element frameworks, would be required.  
Secondly, optimizations such as generalized loop-invariant code motion increase 
the memory requirements due to the introduction of temporary arrays.  This may 
actually lead to poor per-thread shared memory and registers allocation. These 
aspects suggest that the problem of executing generic, automatically-generated 
local assembly kernels on GPUs require a separate, in-depth study. However, in 
the new Section 6, we provide some intuitions of how we could re-adapt COFFEE's 
transformations for execution on GPUs.  Therefore, we invite once more to go 
through the new Section 6 carefully.

@@

******************************************************************************

R3:
We answer here the 5 points of the review.
1) For what concerns execution on a GPGPU, we replicate here the same answer to 
R2.

In some points in the paper (introduction and related work) we cite studies 
concerning local assembly and GPUs. These works however deeply differ from ours 
for a number of reasons, including the facts that they do not rely on automated 
code generation, they are limited to only very low-order methods (typically 
only polynomial order 1), and they do not try to optimize for cross-loop 
arithmetic intensity as we do. The goal in these works is rather to understand 
how to effectively parallelize the local assembly code (so as to have multiple 
threads executing a single kernel, as opposed to our execution model in which 
we assume a single thread per kernel) in specific equations (not in a possibly 
infinite range as we do). A critical issue when executing local assembly 
kernels on GPUs is represented by memory requirements. We give here two 
examples of this problem, although many more issues could be found. First, the 
transfer of the mesh from host to device may be a great obstacle: if too big, 
the mesh
will not fit the available GPU memory, and distributed execution, which is 
marginally supported in available finite element frameworks, would be required.  
Secondly, optimizations such as generalized loop-invariant code motion increase 
the memory requirements due to the introduction of temporary arrays.  This may 
actually lead to poor per-thread shared memory and registers allocation. These 
aspects suggest that the problem of executing generic, automatically-generated 
local assembly kernels on GPUs require a separate, in-depth study. However, in 
the new Section 6 of the revised paper, we provide some intuitions of how we 
could re-adapt COFFEE's transformations for execution on GPUs.

2) We agree that profile-guided optimization could be used to improve the 
performance of kernels, although this aspect is orthogonal to most of our code
transformations, because in no way profile information could suggest, for 
example, to trigger generalized loop-invariant code motion and avoidance of 
redundant computation (described in Section 3.2). Profile-guided optimization 
could theoretically be used for padding and data-alignment of arrays, provided 
that the compiler has sufficient machinery in place to perform (we quote the 
revised paper, Section 6) "some sort of data-flow analysis (...) to ensure that 
the extra loop iterations over the padded region do not affect the numerical 
results.". We discuss on the generality of our code transformation and on their 
integrability in a general-purpose compiler (and in its profile-guided 
optimization system) in the new Section 6 of the paper, which also represents 
novel contribution of our work.

@@

3) TODO

4) Firedrake is *not* specifically designed for Intel architectures and 
compilers. The decision of using Intel platforms and tools is twofold: from one 
hand, we believe that Intel's is one of the most powerful vectorizing and 
optimizing compilers, so for a research like ours it was important to compare 
to the best available on the market; from the other hand, their wide 
availability represents a strong incentive because users will be likely to run 
their experiments on these machines.
It is also worth noting that all general-purpose compilers are, to different 
extents, limited in optimizing for register locality and code motion (for 
instance, this is true for the Intel and GNU compilers, although we only used 
the latter for "local" experimentation), so our work is expected to be 
equally/similarly effective on different yet "standard" CPU architectures.
From the practical viewpoint, we observe that COFFEE introduces optimizations 
aimed at improving cross-loop arithmetic intensity by performing source-level 
transformations, so there are absolutely no technical issues for using our 
domain-specific compiler on architectures different than Intel. Actually, this 
fact is a strength of our work, because it allows COFFEE to be almost 
completely platform-independent.  The only exception regards intrinsics code: 
COFFEE is currently only capable of outputing AVX intrinsics, so at the moment 
we cannot apply vector-register tiling (the also called outer-product 
vectorization technique) to architectures different than x86+AVX and x86+AVX2. 
However, COFFEE's internal algorithms use an intermediate representation of 
intrinsics code, so extending our compiler to other instruction sets would be 
just a matter of adding a new dictionary of intrinsics functions.

5) We agree that expression split has many issues involved. Our cost model 
offers a simple alternative for driving expression splitting. Simplicity is the 
key to effectiveness, portability, and robustness. Probably using a graph 
partitioning approach as you rightly point out would allow us to obtain better 
performance gains, but three points are worth to be highlighted:
- The split factor is a parameter of the transformation. In our experiments we 
  have tested several split factors, which means testing as many code variants 
(i.e. as many as the ways an expression can be split based on sum's associativity) 
and, as shown by the various graphs in the paper, our cost model seems to be 
capable of properly approximating such a parameter. This is of course 
encouraging.
- In the perspective of integrating expression splitting in a general-purpose 
  compiler, or generalizing this technique to other problems (as elaborated in 
the new Section 6), we feel that a simple model is preferable to keep the 
implementation complexity as low as possible.
- Graph partitioning for expression splitting seems to us a complete new 
  research problem. On the other hand, this transformation is only one of the 
many applicable by COFFEE, and we just do not want the focus of the paper being 
on "how to optimally and rapidly split a mathematical expression". Rather, we 
prefer to show that, for our problems, there is clearly a problem of register pressure 
and that by rescheduling the computation in different loops (and 
sub-expressions) we can actually obtain a performance gain. We have a strategy 
for expression splitting; we are not claiming it is the best strategy, but at 
least we demonstrate it is effective and easily predictable.

### 2014-09-08 TACO-paper-v2/coffee.tex
Introduction: copy editing

In many fields such as computational fluid dynamics, computational
electromagnetics and structural mechanics, phenomena are modelled by
partial differential equations (PDEs). Numerical techniques, like the
finite volume method and the finite element method, are widely
employed to approximate solutions of these PDEs. Unstructured meshes
are often used to discretize the computational domain, since they
allow an accurate representation of complex geometries. The
solution is sought by applying suitable numerical operations, or
kernels, to the entities of a mesh, such as edges, vertices, or
cells. On standard clusters of multicores, typically, a kernel is
executed sequentially by a thread, while parallelism is achieved by
partitioning the mesh and assigning each partition to a different node
or thread. Such an execution model, with minor variations, is adopted,
for example, in \cite{pyop2isc}, \cite{Fenics},
\cite{fluidity_manual_v4}, \cite{lizst}.

The time required to apply the numerical kernels is a major issue,
since the equation domain needs to be discretized into an extremely
large number of cells to obtain a satisfactory approximation
of the PDE, possibly of the order of trillions, as in
\cite{Rossinelli2013}. For example, it has been well established that mesh
resolution is critical in the accuracy of numerical weather
forecasts. However, operational forecast centers have a strict time
limit in which to produce a forecast - 60 minutes in the case of the
UK Met Office. Producing efficient kernels has a direct scientific
payoff in higher resolution, and therefore more accurate,
forecasts. Computational cost is a dominant problem in
computational science simulations, especially for those based on
finite elements, which are the subject of this paper.

@@

During the assembly phase, the solution of the PDE is approximated by
executing a problem-specific kernel over all cells, or elements, in the
discretized domain. In this work, we focus on relatively low order finite
element methods, in which an assembly kernel's working set is usually small
enough to fit the L1 cache. Low order methods are by no means exotic: they
are employed in a wide variety of fields, including climate and ocean
modeling, computational fluid dynamics, and structural mechanics. The
efficient assembly of high order methods such as the spectral element
method \cite{SPENCER} requires a significantly different loop nest structure.
High order methods  therefore are excluded from our study.

An assembly kernel is characterized by the presence of an affine,
often non-perfect loop nest, in which individual loops are rather
small: their trip count rarely exceeds 30, and may be as low as 3 for
low order methods. In the innermost loop, a problem-specific,
compute intensive expression evaluates a two dimensional array,
representing the result of local assembly in an element of the
discretized domain. With such a kernel structure, we focus on aspects
like minimization of floating-point operations, register allocation
and instruction-level parallelism, especially in the form of SIMD
vectorization.

Achieving high-performance is non-trivial. The complexity of the
mathematical expressions, often characterized by a large number of
operations on constants and small matrices, makes it hard to determine
a single or specific sequence of transformations that is successfully
applicable to all problems. Loop trip counts are typically small and can
vary significantly, which further exacerbates the issue. A
compiler-based approach is, therefore, the only reasonable option to
obtain close-to-peak performance in a wide range of different local
assembly kernels. Optimizations like padding, generalized
loop-invariant code motion, vector-register tiling, and expression
splitting, as well as their composition, are essential, but very limited is
the support of state-of-the-art polyhedral and vendor compilers, if any. BLAS
routines could theoretically be employed, although a fairly
complicated control and data flow analysis would be required to
automate identification and extraction of matrix-matrix multiplies. In
addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small
dimension of the matrices involved and the potential loss in data
locality can limit or eliminate the performance gain of this approach.

@@

Our experiments show that the original generated code for non-trivial
assembly kernels, despite following state-of-the-art techniques,
% FIXME: what are these state-of-the-art techniques?
remains suboptimal in the context of modern multicore
architectures. Our domain-aware cost-model-driven sequence of code
transformations, aimed at improving SIMD vectorization and register
data locality, can result in performance improvements up to
4.4$\times$ over original kernels, and around 70$\%$ of the test cases
obtain a speed up greater than 2$\times$.

Summarizing, the contributions of this paper are
\begin{enumerate}
\item An optimization strategy for finite element local assembly that
  exploits domain knowledge and goes beyond the limits of both vendor
  and research compilers.
\item The design and implementation of a compiler that automates the
  proposed code transformations for any problems expressible in
  Firedrake.
\item A systematic analysis using a suite of examples of real-world
  importance that is evidence of significant performance improvements on
  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
 \item A full-application evaluation to demonstrate that the optimization
 strategy greatly impacts the whole finite element method.
 \newcounter{enumTemp}
 \setcounter{enumTemp}{\theenumi}
\end{enumerate}
Our contributions also include a theoretical study involving
\begin{enumerate}
 \setcounter{enumi}{\theenumTemp}
 \item An analysis of the generality of the proposed code transformations
 and an investigation of their applicability to different computational
 domains.  
\end{enumerate}
The paper is organized as follows. In Section~\ref{sec:background} we
provide some background on local assembly, show code generated by
Firedrake and emphasize the critical computational
aspects. Section~\ref{sec:code-transf} describes the various code
transformations, highlighting when and how domain knowledge has been
exploited. The design and implementation of our compiler is discussed
in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
shows performance results. The generality of our optimization strategy
and the applicability to other domains is discussed in Section~\ref{sec:generality}. 
Related work is described in Section~\ref{sec:related-work}. Finally, 
Section~\ref{sec:conclusions} reviews our contributions in the light 
of our results, and identifies priorities for future work.

@@

\begin{figure}[h]
\centerline{\includegraphics[scale=0.8]{Pictures/vect-by-vect-inline.pdf}}
\caption{Outer-product vectorization by permuting values in a vector
  register.}
\label{fig:vect-by-vect}
\end{figure}

@@

\begin{figure}[h]
\centerline{\includegraphics[scale=0.8]{Pictures/vect-restore-inline.pdf}}
\caption{Restoring the storage layout after \emph{op-vect}. The figure
  shows how 4$\times$4 elements in the top-left block of the element
  matrix $A$ can be moved to their correct positions. Each rotation,
  represented by a group of three same-colored arrows, is implemented
  by a single shuffle intrinsic.}
\label{fig:restore-layout}
\end{figure}

@@

Splitting an expression (henceforth \emph{split}) has, however, several drawbacks. Firstly, it
increases the number of accesses to $A$ proportionally to the ``split
factor'', which is the number of sub-expressions produced. Also,
depending on how the split is executed, it can lead to redundant
computation. For example, the number of times the product $det*W3[i]$ is
performed is proportional to the number of sub-expressions, as shown in
the code snippet. Further, it increases loop overhead, for example
through additional branch instructions. Finally, it might affect
register locality: for instance, the same array could be accessed in
different sub-expressions, requiring a proportional number of loads be
performed. This is not the case for the Helmholtz example. Nevertheless,
as shown in Section~\ref{sec:perf-results}, the performance gain from
improved register reuse along inner dimensions can still be greater,
especially if the split factor and the splitting itself use heuristics
to minimize the aforementioned issues.

\begin{table}[h]
\tbl{Overview of code transformations for Firedrake-generated assembly kernels.}{
\begin{tabulary}{1.0\columnwidth}{C|C}
\hline
Name (Abbreviation) & Parameter \\\hline
Generalized loop-invariant code motion (\emph{licm}) &   \\
Padding &  \\
Data Alignment & \\
Loop interchange      & loops  \\
Loop unrolling  & unroll factor \\
Outer-product vectorization (\emph{op-vect}) & tile size \\
Expression splitting (\emph{split}) & split point, split factor \\ \hline
\end{tabulary}
}
\label{table:code-transformations}
\end{table}

Table~\ref{table:code-transformations} summarizes the code
transformations described so far. Given that many of these
transformations depend on some parameters, we need a
mechanism to prune such a large optimization space, which can be as large as order of thousands variants for medium-size loops and moderately complex mathematical expressions (all parametric transformations can indeed be combined). This aspect is treated in Section~\ref{sec:pyop2-compiler}.

@@

@@

\section{Related Work}
\label{sec:related-work}
The finite element method is extensively used to approximate solutions
of PDEs. Well-known frameworks and applications include
nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics},
Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake. Numerical
integration is usually employed to implement the local assembly
phase. The recent introduction of domain specific languages (DSLs) to
decouple the finite element specification from its underlying
implementation facilitated, however, the development of novel
approaches. Methods based on tensor contraction~\cite{FFC-Compiler}
and symbolic manipulation~\cite{Francis} have been implemented. We
have designed COFFEE to specifically target numerical integration
because it has been demonstrated that it remains the optimal choice
for a wide class of problems~\cite{quadrature1}.

Optimization of local assembly using numerical integration for CPU
platforms has been addressed in FEniCS~\cite{quadrature1}. The
comparison between COFFEE and this work is presented in
Section~\ref{sec:perf-results-fenics}. In~\cite{Markall20101815}, and
more recently in~\cite{petsc-integration-gpu}, the problem has been
studied for GPU architectures. In~\cite{assembly-opencl}, variants of
the standard numerical integration algorithm have been specialized and
evaluated for the PowerXCell processor, but an exhaustive study from
the compiler viewpoint - like ours - is missing, and none of the
optimizations presented in Section~\ref{sec:code-transf} are
mentioned. Among these efforts, to the best of our knowledge, COFFEE
is the first work targeting low-level optimizations through a real
compiler approach.

>>>>

In many fields such as computational fluid dynamics, computational
electromagnetics and structural mechanics, phenomena are modelled by
partial differential equations (PDEs). Numerical techniques, like the
finite volume method and the finite element method, are widely employed
to approximate solutions of these PDEs. Unstructured meshes are often
used to discretize the computational domain, since they allow an
accurate representation of complex geometries. The solution is sought by
applying suitable numerical operations, described by computational
kernels, to all entities of a mesh, such as edges, vertices, or cells.
On standard clusters of multicores, typically, a kernel is executed
sequentially by a thread, while parallelism is achieved by partitioning
the mesh and assigning each partition to a different node or thread.
Such an execution model, with minor variations, is adopted, for example,
by \citeN{pyop2isc}, \citeN{Fenics}, \citeN{fluidity_manual_v4},
\citeN{lizst}.

The time required to apply the numerical kernels is a major issue,
since the equation domain needs to be discretized into an extremely
large number of cells to obtain a satisfactory approximation
of the PDE, possibly of the order of trillions, as in
\citeN{Rossinelli2013}. For example, it has been well established that mesh
resolution is critical in the accuracy of numerical weather
forecasts. However, operational forecast centers have a strict time
limit in which to produce a forecast - 60 minutes in the case of the
UK Met Office. Producing efficient kernels has a direct scientific
payoff in higher resolution, and therefore more accurate,
forecasts. Computational cost is a dominant problem in
computational science simulations, especially for those based on
finite elements, which are the subject of this paper.

% FIXME: citation for the 30-60% figure

@@

During the assembly phase, the solution of the PDE is approximated by
executing a problem-specific kernel over all cells, or elements, in the
discretized domain. In this work, we focus on relatively low order finite
element methods, in which an assembly kernel's working set is usually small
enough to fit the L1 cache. Low order methods are by no means exotic: they
are employed in a wide variety of fields, including climate and ocean
modeling, computational fluid dynamics, and structural mechanics. The
efficient assembly of high order methods such as the spectral element
method \cite{SPENCER} requires a significantly different loop nest structure.
High order methods are therefore excluded from our study.

An assembly kernel is characterized by the presence of an affine,
often non-perfect loop nest, in which individual loops are rather
small: their trip count rarely exceeds 30, and may be as low as 3 for
low order methods. In the innermost loop, a problem-specific,
compute intensive expression evaluates a two dimensional array,
representing the result of local assembly in an element of the
discretized domain. With such a kernel structure, we focus on aspects
like the minimization of floating-point operations, register allocation
and instruction-level parallelism, especially in the form of SIMD
vectorization.

Achieving high performance is non-trivial. The complexity of the
mathematical expressions, often characterized by a large number of
operations on constants and small matrices, makes it hard to determine a
single or specific sequence of transformations that is successfully
applicable to all problems. Loop trip counts are typically small and can
vary significantly, which further exacerbates the issue. A
compiler-based approach is, therefore, the only reasonable option to
obtain close-to-peak performance in a wide range of different local
assembly kernels. Optimizations like padding, generalized loop-invariant
code motion, vector-register tiling, and expression splitting, as well
as their composition, are essential, but their support in
state-of-the-art polyhedral and vendor compilers, if any, is very
limited. BLAS routines could theoretically be employed, although a
fairly complicated control and data flow analysis would be required to
automate identification and extraction of matrix-matrix multiplies. In
addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small
dimension of the matrices involved and the potential loss in data
locality can limit or eliminate the performance gain of this approach.

@@

Our experiments show that the original generated code for non-trivial
assembly kernels, despite following state-of-the-art techniques,
% FIXME: what are these state-of-the-art techniques?
remains suboptimal in the context of modern multicore
architectures. Our domain-aware cost-model-driven sequence of code
transformations, aimed at improving SIMD vectorization and register
data locality, can result in performance improvements up to
4.4$\times$ over original kernels. Around 70$\%$ of the test cases
obtain a speed up greater than 2$\times$.

Summarizing, the contributions of this paper are:
\begin{enumerate}
\item An optimization strategy for finite element local assembly that
  exploits domain knowledge and goes beyond the limits of both vendor
  and research compilers.
\item The design and implementation of a compiler that automates the
  proposed code transformations for any problems expressible in
  Firedrake.
\item A systematic analysis using a suite of examples of real-world
  importance that is evidence of significant performance improvements on
  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
 \item A full-application evaluation to demonstrate that the optimization
 strategy greatly impacts the whole finite element problem.
 \newcounter{enumTemp}
 \setcounter{enumTemp}{\theenumi}
\end{enumerate}
Our contributions also include a theoretical study involving
\begin{enumerate}
 \setcounter{enumi}{\theenumTemp}
 \item An analysis of the generality of the proposed code transformations
 and an investigation of their applicability to different computational
 domains.
\end{enumerate}
The paper is organized as follows. In Section~\ref{sec:background} we
provide some background on local assembly, show code generated by
Firedrake and emphasize the critical computational
aspects. Section~\ref{sec:code-transf} describes the various code
transformations, highlighting when and how domain knowledge has been
exploited. The design and implementation of our compiler is discussed
in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
shows performance results. The generality of our optimization strategy
and the applicability to other domains is discussed in Section~\ref{sec:generality}.
Related work is described in Section~\ref{sec:related-work}. Finally,
Section~\ref{sec:conclusions} reviews our contributions in the light
of our results, and identifies priorities for future work.

@@

\begin{figure}[h]
\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect-inline.pdf}}
\caption{Outer-product vectorization by permuting values in a vector
  register.}
\label{fig:vect-by-vect}
\end{figure}

@@

\begin{figure}[h]
\centerline{\includegraphics[scale=0.6]{Pictures/vect-restore-inline.pdf}}
\caption{Restoring the storage layout after \emph{op-vect}. The figure
  shows how 4$\times$4 elements in the top-left block of the element
  matrix $A$ can be moved to their correct positions. Each rotation,
  represented by a group of three same-colored arrows, is implemented
  by a single shuffle intrinsic.}
\label{fig:restore-layout}
\end{figure}

@@

Splitting an expression (henceforth \emph{split}) has, however, several drawbacks. Firstly, it
increases the number of accesses to $A$ proportionally to the ``split
factor'', which is the number of sub-expressions produced. Also,
depending on how the split is executed, it can lead to redundant
computation. For example, the number of times the product $det*W3[i]$ is
performed is proportional to the number of sub-expressions, as shown in
the code snippet. Further, it increases loop overhead, for example
through additional branch instructions. Finally, it might affect
register locality: for instance, the same array could be accessed in
different sub-expressions, requiring a proportional number of loads be
performed. This is not the case for the Helmholtz example. Nevertheless,
as shown in Section~\ref{sec:perf-results}, the performance gain from
improved register reuse along inner dimensions can still be greater,
especially if the split factor and the splitting itself use heuristics
to minimize the aforementioned issues.
~\\

Some of the transformations presented in this section, as well as other typical compiler optimizations applicable by COFFEE, like loop interchange and unrolling, are parametric; for instance, expression splitting and loop unrolling depend on the split and the unroll factors, respectively. Since all transformations are composable, the optimization space can be as large as order of thousands variants for local assembly kernels in which mathematical expressions are moderately complex and the iteration space is not particularly small, so a strategy to prune it is required. This problem is treated in the following section.

@@

\subsubsection{On the compilation time}
\label{sec:compile-time}
Firedrake and its various software components (see Figure~\ref{fig:coffee-pipeline}) are implemented in Python and Cython, whereas the generated code is pure C. COFFEE has been written in Python to be naturally integrated within Firedrake. Among the various Firedrake modules, COFFEE is where, in general, the least amount of time is spent. The process of transforming the abstract syntax tree usually takes order of milliseconds. For instance, for the Helmholtz equation with $p=4$ on a tiny tetrahedral mesh, it took 0.0009s to transform the abstract syntax tree, while the whole assembly process needed 0.384s. Being a domain-specific compiler, COFFEE expects C code with the structure outlined in Figure~\ref{code:general-structure}, so typical compilers functionalities such as dependency tracking are simplified with respect to a general-purpose compiler, which implies a clear advantage in compilation time and implementation complexity.

@@

\section{Related Work}
\label{sec:related-work}
The finite element method is extensively used to approximate solutions
of PDEs. Well-known frameworks and applications include
nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics},
Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake. Numerical
integration based on quadrature, as in Firedrake, is usually employed to implement the local assembly
phase. The recent introduction of domain specific languages (DSLs) to
decouple the finite element specification from its underlying
implementation facilitated, however, the development of novel
approaches. Methods based on tensor contraction~\cite{FFC-Compiler}
and symbolic manipulation~\cite{Francis} have been implemented. Nevertheless, it has been demonstrated that quadrature-based integration remains the most efficient choice for a wide class of problems~\cite{quadrature1}, which motivates our work in COFFEE.

Optimization of quadrature-based local assembly for CPU
architectures has been addressed in FEniCS~\cite{quadrature1}. The
comparison between COFFEE and this work has been presented in
Section~\ref{sec:perf-results-fenics}. In~\cite{Markall20101815}, and
more recently in~\cite{petsc-integration-gpu}, the same problem has been
studied for GPU architectures. In~\cite{assembly-opencl}, variants of
the standard numerical integration algorithm have been specialized and
evaluated for the PowerXCell processor, but an exhaustive study from
the compiler viewpoint - like ours - is missing, and none of the
optimizations presented in Section~\ref{sec:code-transf} are
mentioned. Among these efforts, to the best of our knowledge, COFFEE
is the first work targeting low-level optimizations through a real
compiler approach.
